Illuminating a Dark Lens : A Type Ia Supernova Magnified by \\ the \changeR{Frontier Fields} Galaxy Cluster Abell 2744

### 2015-05-21 snTomas.tex
updated with comments from D.Coe; substantial rewrite of abstract

\title{Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by the Galaxy Cluster Abell 2744}

@@

\begin{abstract}
\changenote{Reduced to $<250$ words.}\\
SN \tomas\ is a Type Ia Supernova (SN) discovered at
$z=1.3457\pm0.0001$ behind the galaxy cluster Abell 2744
($z=0.308$). This SN 
%has a projected separation from the cluster core
%of $\sim$40\arcsec, 
is closer to the critical lensing line than any
previous cluster-lensed Type Ia SN, 
%As such, it provides 
providing a valuable
opportunity to test galaxy cluster lens models
with the direct measurement of an absolute magnification at the edge
of the strong-lensing region.  
%We derive a tightly constrained measure
%of the peak apparent magnitude, corrected for light curve shape and
%extinction.  
In a cosmology-independent analysis, we find that \tomas\ is
$0.77\pm0.15$ magnitudes brighter than unlensed Type Ia SNe at similar
redshift, implying a lensing magnification of 
$\mu_{\rm obs}=2.03\pm0.29$. 
\change{In future lens models of Abell 2744, this could
provide some leverage for breaking degeneracies that positional
constraints cannot disentangle.}  We compare the measured $\mu$
against predicted magnifications from 17 lens models, \change{ of
which 13 were generated before the SN magnification was known,
qualifying as a pure ``blind test.''}  The models are collectively
fairly accurate: 8 of the models deliver median magnifications that
are consistent with the measured $\mu$ to within 1$\sigma$.  However,
there is a subtle systematic bias: all of the models that disagree
with the SN $\mu$
%by more than $1\sigma$ 
are {\it higher} than the measured value and \changeR{8} of these
are $\geq1.5\sigma$ above the measured magnification.  We
evaluate possible causes for this mild bias,
%and reject explanations
%that invoke misinterpretation of the SN data or errors in the
%definition of the lens models. We 
and \changeR{find that model accuracy can be improved to some extent
by using multiply-imaged systems vetted with stringent quality cuts
and requiring that a large fraction of those systems have
spectroscopic redshifts.}
% -- at least for this
%cluster and this line of sight -- 
% A more robust evaluation
%\change{Further examination} of systematic biases 
%could be achieved 
%\change{could be pursued with more}
%lens model comparisons using lensed
%Type Ia SNe, as well as cluster lensing simulations.
\end{abstract}

%\keywords{ supernovae: general, supernovae: individual: HFF14Tom, 
%galaxies: clusters: general, galaxies: clusters: individual: Abell 2744, 
%gravitational lensing: strong, gravitational lensing: weak  }

@@

These models represent a broad sampling of the
techniques and assumptions that can be applied to the modeling of mass
distributions in galaxy clusters. 
\change{We discuss some of the modeling choices here, but
a more rigorous comparison of these disparate lens modeling techniques
is beyond the scope of this work. For a complete discussion of each
model's methodology the reader is directed to the listed references.}
Eight of these are so-called {\it parametric}\footnote{Although this
nomenclature is becoming standard in the literature, it is somewhat
misleading, as the pixels or grid cells in free-form models are
effectively parameters as well. Perhaps ``simply-parameterized'' would
be more accurate, though we adopt the more common usage here.}
models, and the remining seven are {\it free-form} models.  
Broadly speaking, the parametric models use parameterized density
distributions to describe the arrangement of mass within the cluster.
%typically associating massive dark matter halos with the positions of
%bright cluster member galaxies. 
Therefore, parametric models rely (to
varying degrees) on the assumption that the cluster's dark matter can
be described by analytic forms such as NFW halos \citep{Navarro:1997},
or pseudo isothermal elliptical mass
distributions \citep[PIEMD][]{Kassiola:1993}.

@@

\change{
A general tendency towards high model magnification can result when
lens models are minimized in the source plane (rather than the image
plane).  For a given model, modifying the surface mass density to have
a shallower profile (declining less quickly with radius) will lead to
an increase in the predicted magnifications across the field.  As the
magnification applies to both flux and area, the shallower profile
also leads to a smaller source plane area, which necessarily brings
delensed images of the same source closer together in the source
plane.  Thus, any models that are optimized in the source plane will
have a tendency toward shallower mass profiles and larger
magnifications, which force all sources toward the same location in
the source plane.  }

\change{  There are several strategies for mitigating this well known bias.
Optimizing a model in the image plane can avoid this bias
entirely \citep{Broadhurst:2005,Zitrin:2009a}.} \changeR{The
Sharon(v2), Zitrin-LTM, Zitrin-NFW, and CATS(v2-2.2) models all use
image plane minimization.}  \change{Alternatively, one may guard against this
bias by using a modified source plane minimization that is constrained
not to generate overly small re-lensed
sources \citep{Sendra:2014,Diego:2014}.} \changeR{In our test set, this
second strategy is employed by the Diego model.}  \change{Finally, the
effect of this bias should be reduced when one uses many strong
lensing constraints over a wide range of redshifts.} \changeR{The
CATS(v2) and CATS(v2.1) models have the largest number of
strong-lensing constraints, using 50 and 55 multiply-imaged systems,
respectively.}

@@

A promising avenue for exploring the origins of systematic biases in
cluster lens models is through the use of simulated lensing data.  One
can start with very deep high-resolution multi-band imaging on an
unlensed field that has a fairly complete spectroscopic redshift
catalog, such as the Hubble Ultra Deep Field.  Then a simulated galaxy
cluster is placed in the field, and the background galaxies are
distorted into arcs and multiple images using a well-defined lensing
prescription.  The artificially lensed images can then be distributed
to lens modeling teams who attempt to reconstruct the (known) mass
profile of the simulated cluster.  This exercise has recently been
pursued with a set of synthetic clusters similar to those observed in
the HFF program (Meneghetti et al. in prep), using techniques
similar to \citet{Meneghetti:2010,Meneghetti:2014}.  Preliminary
analysis of this simulation comparison suggests that
magnifications can be systematically overestimated for
sources that lie outside the strong-lensing region along the minor
axis direction of a simulated cluster that is very
elongated, similar to Abell 2744.  This indicates that some
lens models significantly underestimate the ellipticity of the mass
distribution on large scales for that simulated cluster, which is
consistent with a recent lens model comparison on 25 real clusters
from the CLASH program \citep{Zitrin:2015}. Our analysis of
SN \tomas\ here indicates that such simulation efforts and
head-to-head lens model comparisons will be an important step for
moving toward precision science with cluster-lensed sources.

Increasing the sample of SNe behind clusters like those in
the HFF program -- with rich lensing constraints and deep imaging -- 
would allow this test to be repeated and refined.  With 10 or 100 such
objects, it would be possible to see whether the SN \tomas\ $\mu$
discrepancy is simply an outlier, or an indication of a more
pernicious systematic error.  Any cluster that has been vetted by
pencil-beam magnification tests using lensed \SNIa\ will be able to
provide a more reliable measure of the magnifications for very high
redshift objects.  Similarly, a broad sample of SNe like \tomas\ would
help to define the preferred lens modeling methodology by highlighting
any models that consistently perform well in \SNIa\ lensing tests.
The ongoing FrontierSN program will discover and follow any more
highly magnified SNe that appear behind the Frontier Field
clusters. Unfortunately, the HFF survey is not designed with high-z
transient discovery as a primary science goal, so the FrontierSN
effort will likely add no more than 1-3 new lensed \SNIa. Further
imaging of strong-lensing clusters with \HST\ or the James Webb Space
Telescope (JWST) could enable a larger sample to be collected,
especially if the filters and cadence are optimized for detection
of \SNIa\ at $z>1$.  Massive clusters such as Abell 2744 will continue
to be attractive as cosmic telescopes, allowing the next generation of
telescopes to reach the faintest objects in the very early universe.
The puzzling bias revealed by SN \tomas\ supports a concerted effort
to improve these lenses with further examination of lensing
systematics through simulations and collection of a larger sample of
magnified SNe.

>>>>

\title{Illuminating a Dark Lens : A Type Ia Supernova Magnified by \\ the \changeR{Frontier Fields} Galaxy Cluster Abell 2744}

@@

\begin{abstract}
\changeG{
SN \tomas\ is a Type Ia Supernova (SN) discovered at
$z=1.3457\pm0.0001$ behind the galaxy cluster Abell 2744 ($z=0.308$).
In a cosmology-independent analysis, we find that \tomas\ is
$0.77\pm0.15$ magnitudes brighter than unlensed Type Ia SNe at similar
redshift, implying a lensing magnification of $\mu_{\rm
obs}=2.03\pm0.29$.  This observed magnification provides a rare
opportunity for a direct empirical test of galaxy cluster lens models.
Here we test 17 lens models, 13 of which were generated before the SN
magnification was known, qualifying as pure ``blind tests''.  The
models are collectively fairly accurate: 8 of the models deliver
median magnifications that are consistent with the measured $\mu$ to
within 1$\sigma$.  However, there is a subtle systematic bias: the
significant disagreements all involve models {\it overpredicting} the
magnification. We evaluate possible causes for this mild bias, and
find no single physical or methodological explanation to account for
it.  We do find that model accuracy can be improved to some extent
with stringent quality cuts on multiply-imaged systems, such as
requiring that a large fraction have spectroscopic redshifts.  In
addition to testing model accuracies as we have done here, Type Ia SN
magnifications could also be used as inputs for future lens models of
Abell 2744 and other clusters, providing valuable constraints in
regions where traditional strong- and weak-lensing information is
unavailable.}
\end{abstract}

\keywords{ supernovae: general, supernovae: individual: HFF14Tom, 
galaxies: clusters: general, galaxies: clusters: individual: Abell 2744, 
gravitational lensing: strong, gravitational lensing: weak  }

@@

These models represent a broad sampling of the
techniques and assumptions that can be applied to the modeling of mass
distributions in galaxy clusters. 
\change{We discuss some of the modeling choices here, but
a more rigorous comparison of these diverse lens modeling techniques
is beyond the scope of this work. For a complete discussion of each
model's methodology the reader is directed to the listed references.}
Eight of these are so-called {\it parametric}\footnote{Although this
nomenclature is becoming standard in the literature, it is somewhat
misleading, as the pixels or grid cells in free-form models are
effectively parameters as well. Perhaps ``simply-parameterized'' would
be more accurate, though we adopt the more common usage here.}
models, and the remining seven are {\it free-form} models.  
Broadly speaking, the parametric models use parameterized density
distributions to describe the arrangement of mass within the cluster.
%typically associating massive dark matter halos with the positions of
%bright cluster member galaxies. 
Therefore, parametric models rely (to
varying degrees) on the assumption that the cluster's dark matter can
be described by analytic forms such as NFW halos \citep{Navarro:1997},
or pseudo isothermal elliptical mass
distributions \citep[PIEMD][]{Kassiola:1993}.

@@

\change{
A general tendency towards high model magnification can result when
lens models are minimized in the source plane (rather than the image
plane).  For a given model, modifying the surface mass density to have
a shallower profile (declining less quickly with radius) will lead to
an increase in the predicted magnifications across the field.  As the
magnification applies to both flux and area, the shallower profile
also leads to a smaller source plane area, which necessarily brings
delensed images of the same source closer together in the source
plane.  \changeR{Thus, models that are optimized in the source plane may
have a tendency} toward shallower mass profiles and larger
magnifications, which force all sources toward the same location in
the source plane.  }

\change{  There are several strategies for mitigating this well known bias.
Optimizing a model in the image plane can avoid this bias
entirely \citep{Broadhurst:2005,Zitrin:2009a}.} \changeR{The
Sharon(v2), Zitrin-LTM, Zitrin-NFW, and CATS(v2-2.2) models all use
image plane minimization.}  \change{Alternatively, one may guard against this
bias by using a modified source plane minimization that is constrained
not to generate overly small de-lensed
sources \citep{Sendra:2014,Diego:2014}.} \changeR{In our test set, this
second strategy is employed by the Diego model.}  \change{Finally, the
effect of this bias should be reduced when one uses many strong
lensing constraints over a wide range of redshifts.} \changeR{The
CATS(v2) and CATS(v2.1) models have the largest number of
strong-lensing constraints, using 50 and 55 multiply-imaged systems,
respectively.}

@@

\changeR{
The general problem of combining predictions from many independent
models has a rich history in the astronomical and statistical
literature \citep[e.g.,][]{Press:1997,Hoeting:1999,Liddle:2009}.
In a recent example, \citet{Dahlen:2013} examined the photometric
redshifts from multiple galaxy SED fitting codes using the same input
data.  They found that the photo-z accuracy and precision can be
substantially improved by combining the outputs using a sum of
probability distributions, a hierarchical Bayesian method, or even a
technique as simple as taking the straight median from all available
models.  It may be that lensing magnifications are similar, in that
they can in general be improved through a simple combination of lens
model predictions.  Such model averaging will not, however, resolve a
{\it systematic} bias that is shared by many models.}

A promising avenue for exploring the origins of such systematic biases in
cluster lens models is through the use of simulated lensing data.  One
can start with very deep high-resolution multi-band imaging on an
unlensed field that has a fairly complete spectroscopic redshift
catalog, such as the Hubble Ultra Deep Field.  Then a simulated galaxy
cluster is placed in the field, and the background galaxies are
distorted into arcs and multiple images using a well-defined lensing
prescription.  The artificially lensed images can then be distributed
to lens modeling teams who attempt to reconstruct the (known) mass
profile of the simulated cluster.  This exercise has recently been
pursued with a set of synthetic clusters similar to those observed in
the HFF program (Meneghetti et al. in prep), using techniques
similar to \citet{Meneghetti:2010,Meneghetti:2014}.  Preliminary
analysis of this simulation comparison suggests that
magnifications can be systematically overestimated for
sources that lie outside the strong-lensing region along the minor
axis direction of a simulated cluster that is very
elongated, similar to Abell 2744.  This indicates that some
lens models significantly underestimate the ellipticity of the mass
distribution on large scales for that simulated cluster, which is
consistent with a recent lens model comparison using 25 real clusters
from the CLASH program \citep{Zitrin:2015}. Our analysis of
SN \tomas\ here indicates that such simulation efforts and
head-to-head lens model comparisons will be an important step for
moving toward precision science with cluster-lensed sources.

Increasing the sample of SNe behind clusters like those in the HFF
program -- with rich lensing constraints and deep imaging -- would
allow the test described here to be repeated and refined.  With 10 or
100 such objects, it would be possible to see whether the SN \tomas\
$\mu$ discrepancy is simply an outlier, or an indication of a more
pernicious systematic error.  Any cluster that has been vetted by
pencil-beam magnification tests using lensed \SNIa\ will be able to
provide a more reliable measure of the magnifications for very high
redshift objects.  Similarly, a broad sample of SNe like \tomas\ would
help to define the preferred lens modeling methodology by highlighting
any models that consistently perform well in \SNIa\ lensing tests.
The ongoing FrontierSN program will discover and follow any more
highly magnified SNe that appear behind the Frontier Field
clusters. Unfortunately, the HFF survey is not designed with high-z
transient discovery as a primary science goal, so the FrontierSN
effort will likely add no more than 1-3 new lensed \SNIa. Further
imaging of strong-lensing clusters with \HST\ or the James Webb Space
Telescope (JWST) could enable a larger sample to be collected,
especially if the filters and cadence are optimized for detection
of \SNIa\ at $z>1$.  Massive clusters such as Abell 2744 will continue
to be attractive as cosmic telescopes, allowing the next generation of
telescopes to reach the faintest objects in the very early universe.
The puzzling bias revealed by SN \tomas\ supports a concerted effort
to improve these lenses with further examination of lensing
systematics through simulations and collection of a larger sample of
magnified SNe.
